
\newcommand\lntitle[1]{{\fontsize{14pt}{14pt}\selectfont #1}}

\begin{center}
\begin{tabular}{c}
ERC Starting Grant 2021 \\
Research Proposal [Part B1] \\[1em]
\lntitle{
Satisfiability Modulo Theories and Automated Theorem Proving:}\\[.3em]
\lntitle{Quest for the Missing Link} \\[1em]
ARFUSE
\end{tabular}
\end{center}

\bigskip
\noindent
\begin{tabular}{ll}
Principal investigator (PI) & Shachar Itzhaky \\
Host institution & Technion, Israel Institute of Technology \\
Proposal duration: & 60 months
\end{tabular}

\bigskip\noindent
\fbox{\parbox{0.95\textwidth}{
\quad Automated reasoning using a multitude of logics constitutes a large chunk of today's software analysis and verification.
It is front and center in any automatic verifier, as well as a core component in automatic test generation and synthesis tools.
At present, there exist two predominant approaches to automated reasoning:
\emph{automated theorem proving} (ATP) and
\emph{satisfiability modulo theory} (SMT).
They have somewhat converged toward a common goal, namely, proving the validity of a logical proposition.
However, each brings its own set of techniques to
the table, techniques that are vastly different from each other, and that have evolved more-or-less separately over the last twenty years.
Since each such technique has its own strengths and weaknesses, some problems are solved easily using one tool, but cannot be solved at all by the other.
Combining techniques from these two disciplines has proven to be very challenging since they have significantly diverged.
This difficulty gives rise to the need to form a \textbf{unifying framework for automated reasoning},
that will be able to describe both ATPs and SMT solvers on top of a common mathematical foundation.
\textbf{Bridging the gap} between the ATP and SMT camps will be a significant driving force toward cross-fertilization of ideas,
and push automated reasoning technology forward toward a new generation of automated tools for proving logical propositions.

\quad This research proposes to yield theories and tools that will be most suited to \textbf{software verification tasks}, which are by far the tasks where such automated reasoning is most needed.
By focusing on logical propositions generated as a result of programming-language semantics and software specifications,
we aspire to break the barrier that currently blocks software verification from scaling up and becoming suitable for real-world applications.
The challenges stem not only from the sizes of computer programs, but also of the complex properties that naturally occur in specifications of all but the most trivial instances:
logical quantification, recursive definitions,
and higher-order propositions.
We identify these key factors and propose methods that would allow solving them by drawing from recent advances in both ATP and SMT.
This would allow solving \textbf{new problems} that, at present, cannot solved individually by any of the existing tools.
}}

\clearpage